# Reshaping Data in Stata {#sec-reshape}

## Introduction

Data can be reshaped from *wide* format to *long* format, and back again. Almost any software that is capable of estimating multilevel models is capable of reshaping data.

Below, I detail the procedure for reshaping data in Stata. Here is a sample of the longitudinal data set used in this document.

These data are in *long* format (see @tbl-datalong).

> Every individual in the data has multiple rows. Every row of the data is a *person-timepoint*.

```{r}

library(haven)

library(pander)

```

```{r}

simulated_multilevel_longitudinal_data <- read_dta("simulate-and-analyze-multilevel-data/simulated_multilevel_longitudinal_data.dta")

```

```{r}
#| label: tbl-reshapelongdata
#| tbl-cap: "Data in Long Format"

pander(head(simulated_multilevel_longitudinal_data))

```

## Reshaping Data From Long To Wide

While it is not often that we want to reshape data from *long* to *wide*, I do so here for illustrative purposes. The Stata command for reshaping the data to *wide* format is:

`reshape wide physical_punishment warmth outcome, i(id) j(t)` 

The data are now in *wide* format (See @tbl-datawide).

> Every individual in the data set has a single row of data. Every row in the data set is an individual.

```{r}

simulated_multilevel_longitudinal_data_WIDE <- read_dta("simulate-and-analyze-multilevel-data/simulated_multilevel_longitudinal_data_WIDE.dta")

```

```{r}
#| label: tbl-reshapewidedata
#| tbl-cap: "Data in Wide Format"

pander(head(simulated_multilevel_longitudinal_data_WIDE))

```

## Reshaping Data From Wide To Long

Usually, we are more interested in reshaping data from *wide* to *long*, and that is what I do now.

The command is: 

`reshape long physical_punishment warmth outcome, i(id) j(t)`

If we use this command, we are back to the original format of the data set. 

```{r}
#| label: tbl-reshapelongdata2
#| tbl-cap: "Data in Long Format"

pander(head(simulated_multilevel_longitudinal_data))

```









